Temporal dynamics of food-web structure in response to biodiversity changes in riverine ecosystems

### 2024-08-09 paper/manuscript.Rmd
typo fixing

---
title: "Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs"
author: Alain Danet (a), Loubna El Madouri (a, c), Maud Mouchet (a), Elisa Thébault (d) and Colin Fontaine (a)
date: "Research report"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

While there are widespread evidence of biodiversity decline, little is known how
food-web disassemble in response. Such knowledge is crucial as food-web
structure determine ecosystem functioning and stability to perturbations.
However, assessing how changes in biodiversity affect changes in food-web
structure require the availability of food-web timeseries and to account for the
differential effects of different biodiversity facets on food-web structure.
Here, we used a standardized monitoring of riverine fish communities over the
last decades (1995-2018) in France, from which we inferred the food-web
structure for each sampling events using diet data and allometric scaling rules.
We estimated the temporal trends in community biomass, species richness, and
food-web structure using hierarchical Bayesian models. We assessed the effects
of community biomass and species richness temporal changes on temporal changes
in food-web structure: connectance, average trophic level and redundancy. We
further assess the generality and robustness of our findings by replicating the
analysis over space and using theoretical simulations with a bioenergetic model.
Our result suggest that food-webs that experienced a decline in community
biomass experienced conjointly a loss of species at upper trophic levels and
conversely. While species richness temporal trends had less clear effects on
food-web structure, mainly due through its indirect effects on community biomass
temporal trends. Both community biomass and species richness declines were
associated with a decrease in redundancy, raising concerns about robustness of
such communities to further perturbations. Our study highlight how biodiversity
changes affect food-web structure, raising concerns about consequences on
ecosystem functioning and robustness.

@@

Biodiversity is experiencing massive declines due to continuously increasing
human pressures, which is expected to have striking consequences on ecosystem
functioning and fragility. It is increasing recognised that biodiversity
changes are widely heterogeneous spatially [@mcgill_fifteen_2015;
@blowes_geography_2019], such as species richness declines and increases can be
related to spatial variations in human pressures [@cardinale_is_2018;
@danet_past_2024]. It is increasing recognised that biodiversity changes alter
community structure [@dornelas_assemblage_2014; @hillebrand_biodiversity_2018;
@danet_past_2024], but so far, one key component of biodiversity, species
interactions, has been mostly ignored by studies assessing temporal changes in
biodiversity, despite the importance of biotic interactions for species
coexistence, ecosystem functioning and stability [@chesson_mechanisms_2000;
@thebault_food-web_2003; @oconnor_general_2017; @danet_species_2021]. A better
understanding of biodiversity changes will then need to account how biodiversity
changes alter the structure of ecological networks.

@@

However, assessing the effects of biodiversity changes on food-web structure
requires the availability of multiple temporal timeseries of biodiversity and
food-webs structure, to disentangle the direct and indirect effects of
biodiversity changes on food-web structure, and to account for theories about
the multiple constraints on food-web assembly. We propose to monitor three
structural properties of food-webs and two common monitored biodiversity facets
(i.e. community biomass and species) that are grounded in food-web assembly,
ecosystem functioning, and stability research areas: connectance (i.e. degree of
generalism), weighted average trophic level (i.e. vertical location of biomass),
and redundancy (i.e. the proportion of redundant links). Their are pieces of
evidence to connect those food-web structural metrics to changes in ecosystem
functioning and robustness that we review before to present the temporal
timeseries that allows to assess them.

@@

```{r}
fig_mat_cap <- paste0("Using sites of stream fishes in French streams that
  have been monitored at least 10 years, 1) we infered the food-web at each site
and each sampling event based on ontogenic food diet information for each
species and predator-prey body size ratio. 2) We computed the temporal trends
and the median of community biomass, species richness, connectance, average
trophic level and evenness of biomass at each site. 3) We then assessed the
relationships among food-web structure descriptors, (a) in time and (b) across space.")
```

@@

We found that species richness scaled positively with community biomass, both in
temporal, spatial and theoretical simulations, in accordance with previous
analysis across space in stream communities [@woods_testing_2020;
@danet_species_2021] but also in grasslands, salt marches and forests
[@hector_plant_1999; @tilman_biodiversity_2014; @wu_relationship_2015;
@li_relationship_2018; but see @grace_integrative_2016]. This results is in
agreement with previous findings showing that food-web diversity can increase
biomass by stimulating primary production and resource uptake
[@wang_biodiversity_2018]. Our results suggest that food-web losing species
richness over time also experience a loss of biomass, while those gaining
species experience a increase in community biomass. This results is coherent the
Biodiversity-Ecosystem Functioning theory [@loreau_biodiversity_2001;
@wang_biodiversity_2018], and suggest that the loss of species richness and
biomass might impair ecosystem functioning.

@@

When considering spatial analysis and model simulations, species richness
effects on food-web structure mostly consistent from the temporal analysis. All
analysis converged to indicate that increases in species richness result in
decreases in connectance, a well documented pattern in empirical food-webs
[@dunne_network_2006; @danet_species_2021]. However, spatial analysis found that
species richness decrease average trophic level, while temporal analysis and
theoretical models found that species richness increases result in higher
average trophic level. This result is also in line with other studies which
found a mismatch between spatial and temporal trends [@adler_contrasting_2007;
@johnson_testing_2008]. This discrepancy in space and time is due to the strong
assumption that variations in space reflect differences in assembly stages of
ecological communities and then mirror their temporal assembly/disassembly
trajectories. Temporal changes in species richness might relate to short-term
dynamics while spatial gradients reflect aggregated dynamics over a longer time
scale including a far number of processes including high species turnover along
spatial gradients. Our results then suggest community biomass changes in space
might be a good surrogate to infer assembly and disassembly of food-web in time,
but that it is probably not always the case for species richness.

Our study demonstrates that redundancy of trophic pathways in food-webs are
positively associated to both community biomass and species richness changes, so
as declines in community biomass and species richness triggers a loss in the
proportion of redundant links in food-webs, while increases in community biomass
and species richness over time results in the increase in the proportion of
redundant links in food-webs. Redundant links are the trophic links that can be
removed with triggering secondary extinctions [@allesina_who_2004;
@bodini_using_2009], such an higher proportion of redundant interactions
increase robustness of food-webs to perturbations [@allesina_functional_2009].
It suggests that declines in community biomass and species richness result in a
decline of food-web robustness to further perturbations.

@@

We selected the sites that have been monitored at least 10 years on the
1995-2018 period, with the same sampling protocol over time. We selected
one sampling a year by site. In each site, we ensured low temporal heterogeneity
in sampling period by selecting sampling that were performed in the same period,
leading to a mean standard deviation of sampled month in each site of three
weeks [@danet_species_2021]. Those steps leads to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet), Table S1).

@@

The model parameters were estimated using INLA (Integrated Nested Laplacian
Approximation) Bayesian inference with default uninformative priors
[@rue_bayesian_2017]. The prior distributions for the fixed intercept and slope were
modeled as a flat normal distribution with a mean of zero ($\mathcal{N}(0,\, 1000)$).
Prior distributions for the Gaussian error $\epsilon_{i|n, t}$ and the random
effects were modelled as a log gamma distribution with
shape and inverse scale parameters ($\mathcal{G}(s, \, \tau) = \mathcal{G}(1, \,
5.10^-5)$), then their respective standard deviation were obtained by
backtransforming the inverse scale parameter ($\sigma = 1 / \sqrt{\tau}$).
Community biomass and species richness were logged transformed.

@@

We assessed the validity of the space-for-time substitution hypothesis by
comparing the relationships estimated among food-web structure in time to the
ones inferred in space. We used the same sites as in the temporal analysis,
but we summarised temporal series of each station by the median value over time
of each facet of food-web structure described above (Fig. \@ref(fig:matmet)).

@@

We modelled the relationships among food-web structure, that is
connectance, average trophic level, the proportion of redundant trophic links, and
community biomass ($dB$) and species richness ($dR$) with Structural Equation
Models. We modelled species
richness as related to community biomass and food-web structure. In
turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient, we added a random effect
of the hydrographic basin on the intercept for the latter one.

@@

We used a bioenergetic food-web model to simulate the effects of community
biomass and species richness changes on food-web structure. This food-web model
lifts one of the limitation of our data, that is it includes the biomass of all
the food-webs from primary producers to top predators. This model allows to
simulate complex communities because its parametrization depends mainly on
species bodymasses. We first generated `r nfw` food-webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food-web topology tipically found in empirical
food-webs [@williams_simple_2000]. We set species bodymasses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and bodymass empirically estimated
[@brose_allometric_2006], such as $x_i = aM_i^b$, with $a = xx$ and $b = -0.25$.

@@

We simulated the food-webs dynamics across a gradient of global carrying capacity (K,
range: `r k_range["min"]`-`r k_range["max"]`). The carrying capacity were
standardized according to the number of producers and interspecific competition
($K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K$) to prevent changes in productivity
due to system size [@ives_stability_2000; @loreau_species_2008]. We ran
simulations for 5000 timesteps. At the end of simulations, we removed dead
species and measured community biomass, species richness and food-web structure.
If the final community contained disconnected producers (i.e. without consumers)
or disconnected consumers (i.e. consumers without preys), we discarded them and
run the simulations again with the same parametrisation. The simulations were
done using the Julia package `EcologicalNetworksDynamics.jl`
[@lajaaiti_ecologicalnetworksdynamicsjl_2024].

>>>>

---
title: "Temporal dynamics of food-web structure in response to biodiversity changes in riverine ecosystems"
author: "Alain Danet (a,b), Loubna El Madouri (a,c), Maud Mouchet (a), Elisa Thébault (d) and Colin Fontaine (a)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

While there are widespread evidence of biodiversity decline, little is known how
food-web disassemble in response. Such knowledge is crucial as food-web
structure determine ecosystem functioning and stability to perturbations.
However, assessing how changes in biodiversity affect changes in food-web
structure require the availability of food-web timeseries and to account for the
differential effects of different biodiversity facets on food-web structure.
Here, we used a standardized monitoring of riverine fish communities over the
last decades (1995-2018) in France, from which we inferred the food-web
structure for each sampling events using diet data and allometric scaling rules.
We estimated the temporal trends in community biomass, species richness, and
food-web structure using hierarchical Bayesian models. We assessed the effects
of community biomass and species richness temporal changes on temporal changes
in food-web structure: connectance, average trophic level and redundancy. We
further assess the generality and robustness of our findings by replicating the
analysis over space and using theoretical simulations with a bioenergetic model.
Our results suggest that food-webs that experienced a decline in community
biomass experienced conjointly a loss of species at upper trophic levels and
conversely. While species richness temporal trends had less clear effects on
food-web structure, mainly due through its indirect effects on community biomass
temporal trends. Both community biomass and species richness declines were
associated with a decrease in redundancy, raising concerns about robustness of
such communities to further perturbations. Our study highlights how biodiversity
changes affect food-web structure, raising concerns about consequences on
ecosystem functioning and robustness.

@@

Biodiversity is experiencing massive declines due to continuously increasing
human pressures, which is expected to have striking consequences on ecosystem
functioning and fragility. It is increasing recognised that biodiversity
changes are widely heterogeneous spatially [@mcgill_fifteen_2015;
@blowes_geography_2019], such as species richness declines and increases can be
related to spatial variations in human pressures [@cardinale_is_2018;
@danet_past_2024]. Biodiversity changes result in altered community structure
[@dornelas_assemblage_2014; @hillebrand_biodiversity_2018; @danet_past_2024],
but so far, one key component of biodiversity, species interactions, has been
mostly ignored by studies assessing temporal changes in biodiversity, despite
the importance of biotic interactions for species coexistence, ecosystem
functioning and stability [@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017; @danet_species_2021]. A better understanding of
biodiversity changes will then need to account how biodiversity changes alter
the structure of ecological networks.

@@

However, assessing the effects of biodiversity changes on food-web structure
requires the availability of multiple temporal timeseries of biodiversity and
food-webs structure, to disentangle the direct and indirect effects of
biodiversity changes on food-web structure, and to account for theories about
the multiple constraints on food-web assembly. We propose to monitor three
structural properties of food-webs and two common monitored biodiversity facets
(i.e. community biomass and species) that are grounded in food-web assembly,
ecosystem functioning, and stability research areas: connectance (i.e. degree of
generalism), weighted average trophic level (i.e. vertical location of biomass),
and redundancy (i.e. the proportion of redundant links). There are pieces of
evidence to connect those food-web structural metrics to changes in ecosystem
functioning and robustness that we review before to present the temporal
timeseries that allows to assess them.

@@

```{r}
fig_mat_cap <- paste0("Using sites of stream fishes in French streams that
  have been monitored at least 10 years, 1) we inferred the food-web at each site
and each sampling event based on ontogenic food diet information for each
species and predator-prey body size ratio. 2) We computed the temporal trends
and the median of community biomass, species richness, connectance, average
trophic level and evenness of biomass at each site. 3) We then assessed the
relationships among food-web structure descriptors, (a) in time and (b) across space.")
```

@@

We found that species richness scaled positively with community biomass, both in
temporal, spatial and theoretical simulations, in accordance with previous
analysis across space in stream communities [@woods_testing_2020;
@danet_species_2021] but also in grasslands, salt marches and forests
[@hector_plant_1999; @tilman_biodiversity_2014; @wu_relationship_2015;
@li_relationship_2018; but see @grace_integrative_2016]. This result is in
agreement with previous findings showing that food-web diversity can increase
biomass by stimulating primary production and resource uptake
[@wang_biodiversity_2018]. Our results suggest that food-web losing species
richness over time also experience a loss of biomass, while those gaining
species experience a increase in community biomass. This result is coherent the
Biodiversity-Ecosystem Functioning theory [@loreau_biodiversity_2001;
@wang_biodiversity_2018], and suggest that the loss of species richness and
biomass might impair ecosystem functioning.

@@

When considering spatial analysis and model simulations, species richness
effects on food-web structure mostly consistent from the temporal analysis. All
analysis converged to indicate that increases in species richness result in
decreases in connectance, a well documented pattern in empirical food-webs
[@dunne_network_2006; @danet_species_2021]. However, spatial analysis found that
species richness decrease average trophic level, while temporal analysis and
theoretical models found that species richness increases result in higher
average trophic level. This result is also in line with other studies which
found a mismatch between spatial and temporal trends [@adler_contrasting_2007;
@johnson_testing_2008]. This discrepancy in space and time is due to the strong
assumption that variations in space reflect differences in assembly stages of
ecological communities and then mirror their temporal assembly/disassembly
trajectories. Temporal changes in species richness might relate to short-term
dynamics while spatial gradients reflect aggregated dynamics over a longer
timescale including a far number of processes including high species turnover
along spatial gradients. Our results then suggest community biomass changes in
space might be a good surrogate to infer assembly and disassembly of food-web in
time, but that it is probably not always the case for species richness.

Our study demonstrates that redundancy of trophic pathways in food-webs are
positively associated to both community biomass and species richness changes, so
as declines in community biomass and species richness triggers a loss in the
proportion of redundant links in food-webs, while increases in community biomass
and species richness over time results in the increase in the proportion of
redundant links in food-webs. Redundant links are the trophic links that can be
removed with triggering secondary extinctions [@allesina_who_2004;
@bodini_using_2009], such a higher proportion of redundant interactions
increase robustness of food-webs to perturbations [@allesina_functional_2009].
It suggests that declines in community biomass and species richness result in a
decline of food-web robustness to further perturbations.

@@

We selected the sites that have been monitored at least 10 years on the
1995-2018 period, with the same sampling protocol over time. We selected
one sampling a year by site. In each site, we ensured low temporal heterogeneity
in sampling period by selecting sampling that were performed in the same period,
leading to a mean standard deviation of sampled month in each site of three
weeks [@danet_species_2021]. Those steps lead to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet), Table S1).

@@

The model parameters were estimated using INLA (Integrated Nested Laplacian
Approximation) Bayesian inference with default uninformative priors
[@rue_bayesian_2017]. The prior distributions for the fixed intercept and slope were
modeled as a flat normal distribution with a mean of zero ($\mathcal{N}(0,\, 1000)$).
Prior distributions for the Gaussian error $\epsilon_{i|n, t}$ and the random
effects were modeled as a log gamma distribution with
shape and inverse scale parameters ($\mathcal{G}(s, \, \tau) = \mathcal{G}(1, \,
5.10^-5)$), then their respective standard deviation were obtained by
backtransforming the inverse scale parameter ($\sigma = 1 / \sqrt{\tau}$).
Community biomass and species richness were logged transformed.

@@

We assessed the validity of the space-for-time substitution hypothesis by
comparing the relationships estimated among food-web structure in time to the
ones inferred in space. We used the same sites as in the temporal analysis,
but we summarized temporal series of each station by the median value over time
of each facet of food-web structure described above (Fig. \@ref(fig:matmet)).

@@

We modeled the relationships among food-web structure, that is connectance,
average trophic level, the proportion of redundant trophic links, and community
biomass ($dB$) and species richness ($dR$) with Structural Equation Models. We
modeled species richness as related to community biomass and food-web structure.
In turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient, we added a random
effect of the hydrographic basin on the intercept for the latter one.

@@

We used a bioenergetic food-web model to simulate the effects of community
biomass and species richness changes on food-web structure. This food-web model
lifts one of the limitation of our data, that is it includes the biomass of all
the food-webs from primary producers to top predators. This model allows
simulating complex communities because its parameterization depends mainly on
species bodymasses. We first generated `r nfw` food-webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food-web topology typically found in empirical
food-webs [@williams_simple_2000]. We set species bodymasses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and bodymass empirically estimated
[@brose_allometric_2006], such as $x_i = -0.314M_i^{-0.25}$.

@@

We simulated the dynamics of the `r nfw` food-webs across a gradient of global carrying capacity (K,
range: `r k_range["min"]`-`r k_range["max"]`). The carrying capacity were
standardized according to the number of producers and interspecific competition
($K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K$) to prevent changes in productivity
due to system size [@ives_stability_2000; @loreau_species_2008]. We ran
simulations for 5000 timesteps. At the end of simulations, we removed dead
species and measured community biomass, species richness and food-web structure.
If the final community contained disconnected producers (i.e. without consumers)
or disconnected consumers (i.e. consumers without preys), we discarded them and
run the simulations again with the same parametrisation. The simulations were
done using the Julia package `EcologicalNetworksDynamics.jl`
[@lajaaiti_ecologicalnetworksdynamicsjl_2024]. Then we computed the community
and food-web metrics at the last timesteps and analyzed the data with the same
SEM than described above.
